SMT-Based Generation of Symbolic Automata \thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2019-03-30 2017-Xudong-SMT/ActaInformatica.tex
Acta Inf. paper: simplified the discussion of BIP architectures

Figure~\ref{schema:ArchFailure:BIP} shows a refined version of the
Failure Monitor architecture used in~\cite{CubETH-case-study}.
%
The architecture consists of 2 \emph{coordinator components}
($\NameCtrl$ and $\NameTimer$, for \emph{Control} and \emph{Timer},
respectively), 5 \emph{dangling ports} ($\PortFail$, $\PortResume$,
$\PortTick$, $\PortReset$ and $\PortAsk$) and 5 connectors.
%
%% Contrary to standard BIP models, architectures comprise one or several
%% \emph{operand} components, whereof only the set of \emph{ports} is
%% given.  Here, the operand component is $\NameOprnd$ and its interface
%% consists of the ports $\PortResume$ and $\PortFail$.
%% The two \emph{coordinator} components\mdash $\NameCtrl$ and
%% $\NameTimer$\mdash are standard
The  \emph{behaviour} of BIP components %% insofar as they also
%% have their 
is specified by finite automata extended with
local data variables.  Transitions of these automata are labelled with
the ports of the corresponding components, Boolean guards and update
functions on local variables.  %% For instance, the loop transition
%% $t_1 \goesto{\PortTick, [t < Max], t:=t+1} t_1$
%% in the $\NameTimer$ component is labelled by the port $\PortTick$. It
%% can be fired only when the current value of the local variable $t$ is
%% lower than $Max$.  Upon firing, this transition increments the value
%% of $t$ by $1$.  When omitted, the default guard (\resp update
%% function) is the constant predicate $\true$ (\resp the $\mathit{skip}$
%% operator).  The constant $\mathrm{Max}$ above, is a parameter of the
%% architecture.

@@

\hl{Application of the Failure Monitor architecture ensures that, whenever
a failure is registered in the operand component, the system will be
reset, unless a resumption is registered within $\mathrm{Max}$ time
units (more details in Section~\ref{section:FailureTimer}).}

@@

Figure~\ref{schema:ArchFailure:pNet} shows a pNet encoding of the
above Failure Monitor architecture.  This encoding is structural: each
coordinator component is encoded as a PLTS; dangling ports are encoded as actions of the hole;
%%  and the operand component as a hole;
connectors of the BIP model are encoded as synchronisation vectors.
Each connector that does not involve triggers is trivially encoded by
a synchronisation vector comprising the same ports ($SV2$, $SV3$ and
$SV4$).  In order to encode the semantics of the connectors involving
triggers, we use the classical approach where non-participation of a
port in an interaction is simulated by an additional loop
transition~\cite{milner83-calculi}.  To this end, we consider an
action algebra involving additional Boolean variables: the effective
transitions carry the value $\true$ (\eg $\cstate{0}
\goesto{\PortFail(\true)} \cstate{1}$), whereas the added loops carry
the value $\false$ (\eg $\cstate{2} \goesto{\PortFail(\false)}
\cstate{2}$).  In the synchronisation vectors, we use these variables
to encode the connector structure as a Boolean predicate.  For
example, $SV0$ encodes the connector discussed above: the predicate
$(b1=b2) \land (b1\lor b2 \Rightarrow b0)$ means that the ``true''
transitions $\NameCtrl.\PortFail$ and $\NameTimer.\PortStart$ can only
fire together ($b1 = b2$) and whenever one of them fires,
$\NameOprnd.\PortFail$ must fire also ($b1 \lor b2 \Rightarrow b0$).
This encoding can be systematically obtained for any hierarchical BIP
connector~\cite{BliSif10-causal-fmsd}.
%
\noteEM{Decide later if this is still useful}Although, for the sake of
brevity, we omit priorities from the encoding, this can be easily
achieved, by introducing another set of additional Boolean variables
for relevant ports~\cite{BarBliu15-offer-scico}.

@@

For this example, the tool (using Algorithm 1) builds 408 open transitions, whereof 396
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 12 open
transitions, is shown in Figure~\ref{schema:resultOA1}.

@@

Failure Monitor enforces \noteEM{NO!!!}1)~the safety property \emph{``the
  system reset never happens, unless asked for by a timeout
  following a failure''}, formalised in CTL by
%
\[
\varphi \land \mathtt{AG} (\PortReset \rightarrow \varphi),
\quad\text{where}\
%
\varphi = \mathtt{A} [\lnot \PortReset\ \mathtt{W}\ \PortAsk]
\land \mathtt{A} [\lnot \PortAsk\ \mathtt{W}\ \PortFail],
\]
%
($\mathtt{W}$ being the \emph{weak until} operator) and 2)~the
liveness property \emph{``a reset will be fired when asked for by
  a timeout''}:
%
$\mathtt{AG}(\PortAsk \rightarrow \mathtt{AF}\ \PortReset)$.
%
The satisfaction of the safety property could be established by
applying symbolic model checking techniques.  However, in this
example, it is obvious by inspection of the open automaton.  The
satisfaction of the liveness property relies on the above observation
that in the state $\ostate{s2,t0}$ only the reset transition involves
$\NameCtrl$.  Therefore, under resonable scheduling assumptions,
$\PortReset$ will always be fired.

The safety
property \emph{``There is always a possibility to reset the system
  after a single failure''} (\ie without
  additional failures having to occur in the meantime) enforced by the
  Failure Monitor architecture can be specified using CTL as
    $\AG
\bigl(\PortFail \rightarrow \EX \EU[]{\lnot \PortFail}{
  \PortReset}\bigr)$.  Additional properties are provided in~\cite{Avocs-RR}.

\subsection{Composing Two FailureTimer BIP Architectures}
\label{section:use-case:Double}

In the conception of the nanosatellite software \cite{CubETH-case-study}, the FailureTimer architecture was used to control several processes
communicating on the common bus. Timeout signals (``$ask$''), resets,
and the passing of time ($tick$) were synchronised in the global
system.
In this section, we use a combination of 2 FailureTimer architectures
to demonstrate how the encoding works, and as a slightly larger
example to show how the approach scales up.

@@

In Figure~\ref{schema:ArchFailure:BIP:Double} we show the combination
of two FailureTimer architectures, with their respective controlled
processes.
In Figure~\ref{schema:ArchFailure:Double:OA} we show the resulting
open automaton. It indicates that the two timers are forced to
be synchronous: each can only tick after the other has also started,
and then they will tick, reset, and timeout, synchronously.

This is a very strong assumption, and we are working on a more
flexible specification (at BIP level) to release this hypothesis while
keeping the safety properties of the composition.

>>>>

Figure~\ref{schema:ArchFailure:BIP} shows a refined version of the
Failure Monitor architecture used in~\cite{CubETH-case-study}.
%
The architecture consists of 2 \emph{coordinator components}
($\NameCtrl$ and $\NameTimer$, for \emph{Control} and \emph{Timer},
respectively), 5 \emph{dangling ports} ($\PortFail$, $\PortResume$,
$\PortTick$, $\PortReset$ and $\PortAsk$) and 5 connectors.
%
%% Contrary to standard BIP models, architectures comprise one or several
%% \emph{operand} components, whereof only the set of \emph{ports} is
%% given.  Here, the operand component is $\NameOprnd$ and its interface
%% consists of the ports $\PortResume$ and $\PortFail$.
%% The two \emph{coordinator} components\mdash $\NameCtrl$ and
%% $\NameTimer$\mdash are standard
The  \emph{behaviour} of BIP components %% insofar as they also
%% have their 
is specified by finite automata extended with
local data variables.  Transitions of these automata are labelled with
the ports of the corresponding components, Boolean guards and update
functions on local variables.  %% For instance, the loop transition
%% $t_1 \goesto{\PortTick, [t < Max], t:=t+1} t_1$
%% in the $\NameTimer$ component is labelled by the port $\PortTick$. It
%% can be fired only when the current value of the local variable $t$ is
%% lower than $Max$.  Upon firing, this transition increments the value
%% of $t$ by $1$.  When omitted, the default guard (\resp update
%% function) is the constant predicate $\true$ (\resp the $\mathit{skip}$
%% operator).  The constant $\mathrm{Max}$ above, is a parameter of the
%% architecture.
Although this is not visible in the figure, below we will assume that
the dangling ports $\PortFail$ and $\PortResume$ belong to the operand
component, whereas the dangling ports $\PortTick$, $\PortAsk$ and
$\PortReset$ represent actions of the environment.

@@

Intuitively, application of the Failure Monitor architecture ensures
that, whenever a failure is registered in the operand component, the
system will be reset, unless a resumption is registered within
$\mathrm{Max}$ time units.  Formally, this statement is decomposed
into several properties, among which 
%
1)~the safety property \emph{``There is always a possibility to reset
  the system after a single failure''} (\ie without additional
failures having to occur in the meantime) specified using CTL as
%
\begin{equation}
  \label{eq:FM:safety}
  \AG
  \bigl(\PortFail \rightarrow
  \EX \EU[]{\lnot \PortFail}{\PortReset}\bigr)
\end{equation}
%
and 2)~the liveness property \emph{``a reset will be fired when asked
  for by a timeout''}:
%
\begin{equation}
  \label{eq:FM:liveness}
  \mathtt{AG}(\PortAsk \rightarrow \mathtt{AF}\ \PortReset).
\end{equation}
%
Additional properties are provided in~\cite{Avocs-RR}.

@@

Figure~\ref{schema:ArchFailure:pNet} shows a pNet encoding of the
above Failure Monitor architecture.  This encoding is structural: each
coordinator component is encoded as a PLTS; dangling ports belonging
to the operand component are encoded as actions of the hole; 
%%  and the operand component as a hole;
connectors of the BIP model are encoded as synchronisation vectors,
\noteSB{Eric, please check whether this is ok.}\hl{with ports
  representing the actions of the environment used as global actions}.
Each connector that does not involve triggers is trivially encoded by
a synchronisation vector comprising the same ports ($SV2$, $SV3$ and
$SV4$).  In order to encode the semantics of the connectors involving
triggers, we use (in $SV0$ and $SV1$) the classical approach where
non-participation of a port in an interaction is simulated by an
additional loop transition~\cite{milner83-calculi}.  To this end, we
consider an action algebra involving additional Boolean variables: the
effective transitions carry the value $\true$ (\eg $\cstate{0}
\goesto{\PortFail(\true)} \cstate{1}$), whereas the added loops carry
the value $\false$ (\eg $\cstate{2} \goesto{\PortFail(\false)}
\cstate{2}$).  In the synchronisation vectors, we use these variables
to encode the connector structure as a Boolean predicate.  For
example, $SV0$ encodes the connector discussed above: the predicate
$(b1=b2) \land (b1\lor b2 \Rightarrow b0)$ means that the ``true''
transitions $\NameCtrl.\PortFail$ and $\NameTimer.\PortStart$ can only
fire together ($b1 = b2$) and whenever one of them fires,
$\NameOprnd.\PortFail$ must fire also ($b1 \lor b2 \Rightarrow b0$).
This encoding can be systematically obtained for any hierarchical BIP
connector~\cite{BliSif10-causal-fmsd}.
%
\noteEM{Decide later if this is still useful}Although, for the sake of
brevity, we omit priorities from the encoding, this can be easily
achieved, by introducing another set of additional Boolean variables
for relevant ports~\cite{BarBliu15-offer-scico}.

@@

For this example, the tool (using Algorithm 1) builds 408 open transitions, whereof 396
are detected unsatisfiable by Z3.  The resulting open automaton, with
3 reachable global states (out of the possible 6) and 12 open
transitions, is shown in Figure~\ref{schema:resultOA1}.\noteSB{Eric, in the state 20, $b0$ should always be $false$: $\PortFail(true)$ is not possible in Control, and the safety property (\ref{eq:FM:safety}) does not hold otherwise.}

@@

The satisfaction of the safety property (\ref{eq:FM:safety}) could be
established by applying symbolic model checking techniques.  However,
in this example, it is obvious by inspection of the open automaton.
The satisfaction of the liveness property relies on the observation
that in the state $\ostate{s2,t0}$ only the reset transition involves
$\NameCtrl$.  Therefore, under resonable scheduling assumptions,
$\PortReset$ will always be fired.

\subsection{Composing Two Failure Monitor BIP Architectures}
\label{section:use-case:Double}

%% In the design of the nanosatellite software \cite{CubETH-case-study}, the Failure Monitor architecture was used to control several processes
%% communicating through a common bus. Timeout signals (``$ask$''), resets,
%% and the passing of time ($tick$) were synchronised in the global
%% system.
In this section, we use a combination of two Failure Monitor
architectures shown in Figure~\ref{schema:ArchFailure:BIP:Double}
%% to demonstrate how the encoding works, and
as a slightly larger example to show how the approach scales up.  In
this example, each of the two instances of the architecture is applied
to a corresponding operand component.  Hence, the dangling ports
$\PortFail$ and $\PortReset$ are duplicated accordingly.  On the
contrary, the environment actions $\PortTick$, $\PortAsk$ and
$\PortReset$ are shared.  For a detailed presentation of the
architecture composition, we refer the reader to
\cite{AttieBBJS16-architectures-faoc}.

@@

%% In Figure~\ref{schema:ArchFailure:BIP:Double} we show the combination
%% of two Failure Monitor architectures, with their respective controlled
%% processes.
In Figure~\ref{schema:ArchFailure:Double:OA} we show the open
automaton computed for the combined architecture of
Figure~\ref{schema:ArchFailure:BIP:Double}.  It indicates that the two
timers are forced to be synchronous: each can only tick after the
other has also started, and then they will tick, reset, and timeout,
synchronously.
%
\hl{This is a very strong assumption, and we are working on a more
flexible specification (at BIP level) to release this hypothesis while
keeping the safety properties of the composition.}

\noteInSB{The idea was that, by analysing the combination, we observe
  that the architecture does not compose well with itself, generating
  a deadlock (or an implicit deadlock, where only $false$ transitions
  canb be fired.  Based on this observation, we were to improve it
  accodingly.  The problem is that I do not see any deadlocks in
  Figure~\ref{schema:ArchFailure:Double:OA}.  I presume that some of
  the asterisks should be only $false$ or something of a kind.}
